Clustering Beat-Chroma Patterns\\ in a Large Music Database

### 2010-03-24 Paper/Paper.tex
lots of tweaks to section 4

\begin{figure}[thb]
\begin{center}
\includegraphics[width=.6\columnwidth]{code}
\end{center}
\caption{\small{A typical codeword from a codebook of size $200$. 
Patterns represented
$2$ bars and the pattern length was set to $8$.
}}
\label{fig:code}
\end{figure}

@@

\section{Experiments}\label{sec:experiments}
In this section we present different clustering experiments. We introduce
our training data and our main testing data. Some detailed settings
of our algorithm are also provided. Then, we answer some common questions
about any clustering algorithms, for instance 
``how the number of codes affect the distortion''
and ``how many learning samples should we use''.
Note that we will present two additional experiments in 
Section \ref{sec:exps2}.

\subsection{Training: Cowbell Dataset}\label{sec:traindata}
We have $43,300$ tracks that were uploaded
to morecowbell.dj\footnote{http://www.morecowbell.dj/}, 
an online service that uses the Echo
Nest to produce a remixed version of
an uploaded music file that adds cowbell and other sound
effects in time with the music.
The $43.3$K songs, giving $\mathbf{3.7}$\textbf{M}
non silent bars, a bar usually
contains $4$ beats (Figure \ref{fig:barsize}).

\subsection{Testing: USPOP}\label{sec:testdata}
We had access to a low quality (32kbps) version of the songs in the uspop 2002 
data set \cite{uspop2002}.
This well-known data show a great diversity of pop songs, but is not particular
otherwise\footnote{We 
also tried to use Tzanetakis dataset, 
%\cite{Tzanetakis2002a},
but the $30$ seconds segments seemed hard to analyze by the Echo Nest API, 
most did not contain any bar.}.
uspop2002 serves as testing set to measure how well a codebook learned on
the Cowbell data set can represent new songs. We get Echo Nest features
for $8651$ songs of that dataset.

\subsection{Setting}\label{ssec:setting}
We take one or two bars and normalize the patches to size 4 or 8 respectively.
We roll the patch to be invariant to the key (on the patch level, not on
the song level). We learn a codebook of size $K$ over the cowbell dataset 
using the VQ algorithm (Algorithm \ref{algo:vq}). A typical learning rate 
is $0.01$ for $200$ iterations over the whole dataset.

We use the codebook to encode our testing set (see Section \ref{sec:testdata}).
Each pattern is encoded with only one code. We can measure the average
distance between a pattern and its encoding. We can also measure the use
of the different codes, \ie the size of the cluster around a code.

As a distance measure (or distortion measure) between patterns, we use
the average squared euclidean distance. In details, if a pattern $p_1$
is composed of elements $p_{1ij}$, and similarly for a pattern $p_2$,
the distance between $p_1$ and $p_2$ is:
\begin{eqnarray}
  \dist(p_1,p_2) = \sum_{i,j} \frac{(p_{1ij} - p_{2ij})^2}{\size(p_1)}
  \label{eq:dist}
\end{eqnarray}
We assume $p_1$ and $p_2$ have the same size.

\subsection{Basic Properties}
This section present some fundamental results of the clustering algorithm.
These are important, but do not change the larger picture. Due to space
constraints, we give few details and refer to the figures.
\begin{itemize}
\item A larger dataset gives a better encoding codebook. See Figure
\ref{fig:data_sizes}.
\item A larger codebook better encodes new songs. See 
Table \ref{tab:cbsize}.
\item Larger patterns are more difficult to encode, thus require
larger codebooks. See Figure \ref{fig:perbeat}.
\end{itemize}

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.99\columnwidth]{data_sizes}
\end{center}
\caption{\small{Distortion for a codebook of size $100$ encoding one bar
at a time, a bar being represented by $4$ chroma feature vector.
Therefore, a codeword or a pattern is of size $12$ x $4$ = $48$.
Distortion measure on the test data set. Data size tested range
from $0$ (just initialization) to $500$ thousand. Patterns were selected at
random from the dataset of approximately $3,7$ million patterns, except
for the dotted line (non random in the legend). In that case we took the
first $N$ samples in our dataset. The difference is only visible for small
training sets ($1$K).
}}
\label{fig:data_sizes}
\end{figure}

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.9\columnwidth]{encoding_per_beat}
\end{center}
\caption{\small{Encoding larger and larger patterns with a fixed size
codebook. The size of the pattern is defined by the number of beats,
the bar information was not taken into account for that experiment.
}}
\label{fig:perbeat}
\end{figure}

@@

We train a codebook of size $200$ over patterns of size $12$x$8$ covering
$2$ bars at a time. Results shown are on a second testing set, see
Subsection \ref{ssec:artist}.

The 25 most frequently used codewords in the test set are shown in
Figure \ref{fig:codes1}.  
The frequency of use of these codewords are shown in Figure \ref{fig:freqs}.
The codewords primarily consist of sustained
notes and simple chords.  Since they are designed to be key-invariant,
specific notes do not appear.  Instead the first 7 codewords
correspond to a single note sustained across two bars (codeword 1),
perfect fifth (codes 2 and 3) and fourth intervals (codes 4 7), and a
major and minor triads (codeword 6 and 5).  Many of the remaining
codewords correspond to common transitions from one chord to another,
\eg a transition from one major triad to another (\ie Cmaj
$\rightarrow$ Fmaj assuming that chroma 0 is C) in codes 8 and 10 and
the reverse transition in code (\ie Fmaj $\rightarrow$ Cmaj) in code
22.  Does this transition have a better name like ``relative x''?

\begin{figure}[h]
\begin{center}
\includegraphics[width=.9\columnwidth]{codes1}
\end{center}
\caption{\small{
Most used codes in the codebook in term of frequency of use in the test
set (artist20, see Subsection \ref{ssec:artist}). 
This codebook was of size $200$ and trained on $2$ bars, pattern
size is $12$x$8$. Code in Figure \ref{fig:code} is the $8$th most used
here (second row, third column). 
Numbers display are the index (from most used to least) and in parentheses
the percentage of use in the whole collection times.
There is $71832$ patterns in total in the dataset.
}}
\label{fig:codes1}
\end{figure}

@@

Finally, more related to our particular data, we try to assess how many
sustained patterns are represented in the codebook. We look at
the average variance over time in Figure \ref{fig:code_var}.
25\% of the codebook correspond
to very simple sustained patterns similar to the top row of Figure
\ref{fig:codes1}. For $2$ bars codewords, it seems that interesting
patterns appear in only $33$\% of the data.

Showing codewords of $2$ bars is practical and consistent with our experiments
in Section \ref{sec:exps2}. However, longer codewors can also reveal
interesting structures. A random samples of such codewords is presented
in Figure \ref{fig:codes8bars}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.9\columnwidth]{codes_lle}
\end{center}
\caption{\small{
Locally linear embedding (LLE) visualization of the codebook.
LLE tries to keep near neighbors (here according to euclidean distance)
in the same $2$D neighborhood. Shown patterns were randomly selected
in each neighborhood.
}
ronw: I think this figure would be better if it only included the
bottom subplot.
}
\label{fig:lle}
\end{figure}

@@

@@

We apply our codebook to the artist recognition task.
Our data set encompass $1402$ songs from $20$ artists, 
mostly rock and pop of different subgenres. Results exist in the
litterature for that set \cite{Ellis2007}. The best accuracy is
$59\%$, whereas using only chroma as a representation yields an
accuracy of $33\%$. The latter results is obtained through a
gaussian mixture model.

@@

As for enhancements, we specifically did not compress our features using
gaussian mixtures or other generative model. That being said, these methods
could be use to develop better distance measures between patterns.
The use of the euclidean distance is arbitrary. Summarizing patches
with gaussians, and then comparing the distance between those gaussians,
could reduce the influence of the noise in the distance measure.

Moving on to larger scales, we could imagine merging codes together and
splitting some on two if the clusters become too small or too large.We could
also cluster the codebook itself, in a similar fashion as hierarchical
gaussian mixtures \cite{Vasconcelos2001}.
%  (similar idea used for indexing \cite{Muja2009})

>>>>

\begin{figure}[thb]
\begin{center}
\includegraphics[width=.6\columnwidth]{code}
\end{center}
\caption{\small{A typical codeword from a codebook of size $200$ (code
    7 in Figure \ref{fig:codes1})
Patterns represented
$2$ bars and the pattern length was set to $8$.
}}
\label{fig:code}
\end{figure}

@@

\section{Experiments}\label{sec:experiments}
In this section we present different clustering experiments. We introduce
our training data and our main testing data. Some detailed settings
of our algorithm are also provided. Then, we answer some common questions
about any clustering algorithms, for instance 
``how the number of codes affect the distortion''
and ``how many learning samples should we use''.
% Note that we will present two additional experiments in 
% Section \ref{sec:exps2}.

\subsection{Data}
%\subsection{Training data: Cowbell Dataset}
\label{sec:traindata}
%\subsection{Test data: USPOP}
\label{sec:testdata}
Our training data consists of $43,300$ tracks that were uploaded to
morecowbell.dj,\footnote{http://www.morecowbell.dj/} an online service
based on the Echo Nest analyze API which remixes uploaded music by
adding cowbell and other sound effects synchronized in time with the
music.  The $43.3$K songs contain $\mathbf{3.7}$ million non silent
bars which we cluster using the approach described in the previous
section.

For testing, we made use of low quality (32kbps) version of the songs
from the uspop2002 data set \cite{uspop2002}.  This well-known data
contains a wide variety of pop songs, but is not particular otherwise.
% \footnote{We also tried to use Tzanetakis dataset,
% %\cite{Tzanetakis2002a},
% but the $30$ seconds segments seemed hard to analyze by the Echo Nest API, 
% most did not contain any bar.}.
uspop2002 serves as test set to measure how well a codebook learned on
the Cowbell data set can represent new songs.  We obtained Echo Nest features
for the $8651$ songs contained in the dataset.

\subsection{Setting}\label{ssec:setting}
We take one or two bars and normalize the patches to size 4 or 8
respectively.  We learn a codebook of size $K$ over the Cowbell
dataset using the VQ algorithm (Algorithm \ref{algo:vq}). We used a
learning rate of $\ell=0.01$ for $200$ iterations over the whole
dataset.
%
We use the codebook to encode the test set.
Each pattern is encoded with only one code. We can measure the average
distance between a pattern and its encoding. We can also measure the use
of the different codes, \ie the size of the cluster around a code.

We use the average squared Euclidean distance as the distortion
measure between chroma patches.  Given a pattern $p_1$
composed of elements $p_{1ij}$, and a similar pattern $p_2$, the
distance between them is:
\begin{eqnarray}
  \dist(p_1,p_2) = \sum_{i,j} \frac{(p_{1ij} - p_{2ij})^2}{\size(p_1)}
  \label{eq:dist}
\end{eqnarray}
We assume $p_1$ and $p_2$ have the same size.  This is enforced by the
meter normalization procedure described in Section \ref{sec:approach}.

\subsection{Basic Properties}
FIXME: flesh this out a \emph{little} more?

This section present some fundamental results of the clustering algorithm.
These are important, but do not change the larger picture. Due to space
constraints, we give few details and refer to the figures.
\begin{itemize}
\item Encoding performance improves with increasing training data. See Figure
\ref{fig:data_sizes}.
\item Encoding performance improves with increasing codebook size. See 
Table \ref{tab:cbsize}.
\item Larger patterns are more difficult to encode, thus requiring
larger codebooks. See Figure \ref{fig:perbeat}.
\end{itemize}

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.99\columnwidth]{data_sizes}
\end{center}
\caption{\small{Distortion for a codebook of size $100$ encoding one bar
at a time, a bar being represented by $4$ chroma feature vector.
Therefore, a codeword or a pattern is of size $12$ x $4$ = $48$.
Distortion measure on the test data set. Data size tested range
from $0$ (just initialization) to $500$ thousand. Patterns were selected at
random from the dataset of approximately $3,7$ million patterns, except
for the dotted line (non random in the legend). In that case we took the
first $N$ samples in our dataset. The difference is only visible for small
training sets ($1$K).
%
FIXME: Can we kill the ``non-random'' line.  It doesn't really add anything.
}}
\label{fig:data_sizes}
\end{figure}

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.9\columnwidth]{encoding_per_beat}
\end{center}
\caption{\small{Encoding larger and larger patterns with a fixed size
codebook. The size of the pattern is defined by the number of beats,
the bar information was not taken into account for this experiment.
}}
\label{fig:perbeat}
\end{figure}

@@

We train a codebook of size $200$ over patterns of size $12$x$8$ covering
$2$ bars at a time.  The results shown are on the test set from
Section \ref{ssec:artist}.

The 25 most frequently used codewords in the test set are shown in
Figure \ref{fig:codes1}.  
The frequency of use of these codewords are shown in Figure \ref{fig:freqs}.
The codewords primarily consist of sustained
notes and simple chords.  Since they are designed to be key-invariant,
specific notes do not appear.  Instead the first 7 codewords
correspond to a single note sustained across two bars (codeword 1),
perfect fifth (codewords 2 and 3) and fourth intervals (codewords 4
and 7), and a
major and minor triads (codewords 6 and 5, respectively).  Many of the remaining
codewords correspond to common transitions from one chord to another,
\eg a transitions between major triads in codes 8 and 10 ( Cmaj
$\rightarrow$ Fmaj)% assuming that chroma 0 is C) 
and the reverse transition in code 22 (\ie Fmaj $\rightarrow$ Cmaj).
FIXME: Does this transition have a better name like ``relative x''?

\begin{figure}[h]
\begin{center}
\includegraphics[width=.9\columnwidth]{codes1}
\end{center}
\caption{\small{
Most frequently used codes in the codebook %in term of frequency of use
in the artist20 test set. 
This codebook contains $200$ codes trained on $2$ bar, $12 \times 8$
patches.
The numbers 
%above each panel correspond to the display are the index (from most used to least) and
in parentheses correspond to 
the percentage of use in the test set.
%There are $71832$ patterns in total in the dataset.
}}
\label{fig:codes1}
\end{figure}

@@

Finally, more related to our particular data, we try to assess how many
sustained patterns are represented in the codebook. We look at
the average variance over time in Figure \ref{fig:code_var}.
25\% of the codebook correspond
to very simple sustained patterns similar to the top row of Figure
\ref{fig:codes1}. For $2$ bars codewords, it seems that patterns
containing more complex temporal activity account for only $33$\% of
the data.

Showing codewords of $2$ bars is practical and consistent with our experiments
in Section \ref{sec:exps2}. However, longer codewords can also reveal
interesting structure. A random sample of such codewords is presented
in Figure \ref{fig:codes8bars}.  FIXME: why a random sample?  Can we
show their usage in the test (or train) set as in fig \ref{fig:codes1}?

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.9\columnwidth]{codes_lle}
\end{center}
\caption{\small{
Locally linear embedding (LLE) visualization of the codebook.
LLE tries to keep near neighbors (here according to euclidean distance)
in the same $2$D neighborhood. Shown patterns were randomly selected
in each neighborhood.
}
FIXME:Can we make this a little bit bigger so the patterns are
actually visible?  Does the figure add much over Fig \ref{fig:codes1}?
Maybe it has some more complex patterns? 
}
\label{fig:lle}
\end{figure}

@@

FIXME: we want the differences to be close to uniformly random, right
(hence centroid)?
And that's kind of what they look like.  Does Fig
\ref{fig:cluster_diff} really add much?

@@

We apply our codebook to the artist recognition task.
Our data set encompass $1402$ songs from $20$ artists, 
mostly rock and pop of different subgenres. Results exist in the
litterature for that set \cite{Ellis2007}. The best accuracy is
$59\%$, whereas using only chroma as a representation yields an
accuracy of $33\%$. The latter results is obtained through a
Gaussian mixture model.

@@

As for enhancements, we specifically did not compress our features using
Gaussian mixtures or other generative model. That being said, these methods
could be use to develop better distance measures between patterns.
The use of the euclidean distance is arbitrary. Summarizing patches
with Gaussians, and then comparing the distance between those Gaussians,
could reduce the influence of the noise in the distance measure.

Moving on to larger scales, we could imagine merging codes together and
splitting some on two if the clusters become too small or too large.We could
also cluster the codebook itself, in a similar fashion as hierarchical
Gaussian mixtures \cite{Vasconcelos2001}.
%  (similar idea used for indexing \cite{Muja2009})
